---
title: "Final Project Proposal"
author: "Moriah Ruggerio"
subtitle: National Park Species
output:
  html_document:
    df_print: paged
  tufte::tufte_html:
    tufte_variant: envisioned
    highlight: pygments
    css: ../lab.css
  pdf_document:
    latex_engine: xelatex
link-citations: true
---

```{r setup, include=FALSE}
#knitr::opts_chunk$set(eval = TRUE)

library(tidyverse)
library(skimr)
library(infer)
```

# Import the data:
```{r}
most_visited_nps_species_data <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2024/2024-10-08/most_visited_nps_species_data.csv')

```

``` {r tidydata}
most_visited_nps_species_data_cleaned <- most_visited_nps_species_data %>%
  janitor::clean_names()
```

Determine what variables to remove to keep it between 10-20:
```{r}

most_visited_nps_species_data_cleaned <- most_visited_nps_species_data_cleaned %>%
  filter(taxon_record_status == "Active", park_accepted == TRUE) %>%  #only includes data with active taxons, and accepted in park status
  #count(taxon_record_status)
  select(park_name:family, sci_name, common_names, occurrence:abundance, references:vouchers, te_status:s_rank) #%>% # removes empty and exluded rows to keep the variables within range
  #count(taxon_record_status, park_accepted)
  
#no data in synonyms, external_links
#all data is not sensitive (remove variable)
#combine nativeness_tabs with native & same with occurrence and occurrence_tags
#remove park_tags
#remove inactive taxons (including data rows)
#only include park accepted/remove the row and record status
```

``` {r}
write_csv(most_visited_nps_species_data_cleaned, file = "data/most_visited_nps_species_data_cleaned.csv")
```

``` {r}
most_visited_nps_species_data_cleaned %>%
  count(nativeness)
```

# Introduction

This project uses the dataset `most_visited_nps_species_data_cleaned`. The National Park Service (NPS) keeps a list of every species found within each national park. The data used in this project is originally from the National Park Species List. It can be accessed here: https://irma.nps.gov/NPSpecies/Search/SpeciesList. Due to the massive size of this dataset, it was restricted to the top 15 most popular national parks by  `f. hull`  in the `tidytuesday` git repository on September 2, 2024. This reduced dataset (`most_visited_nps_species_data`) was accessed on March 15, 2025 from https://github.com/rfordatascience/tidytuesday/blob/main/data/2024/2024-10-08/readme.md. Since this dataset originally had too many for the confines of this assignment, I restricted the scope of the data further. Only included in this projects' dataset (`most_visited_nps_species_data_cleaned`)  are species whose presence in the park has been approved by the NPS, have an active taxon (current scientific name), and a non-sensitive species status. Data on sensitive species is removed from all public access by the NPS to protect endangered or fragile species in national parks. 

The dataset has `r nrow(most_visited_nps_species_data_cleaned)` rows and `r ncol(most_visited_nps_species_data_cleaned)` variables. Each row is a specific species found in one of the parks. The variables are `park_name`, `category_name`, `order`, `family`, `sci_name`, `common_names`, `occurrence`, `occurrence_tags`, `nativeness`, `nativeness_tags`, `abundance`, `references`, `observations`, `vouchers`, `te_status`, `state_status`, `ozone_sensitive_status`, `g_rank`, `s_rank`. (See `README.md` for variable descriptions. The intent of this research project is to examine if there is trends between biodiversity and the popularity of a park. Some specific research questions to be examined are: (1) Do more popular parks have higher observations of birds (and other classes)? (2)  Are there more native species seen in popular parks than invasives or introduced?

# Data

```{r}
glimpse(most_visited_nps_species_data_cleaned)

skim(most_visited_nps_species_data_cleaned)
```

# Data Analysis Plan

*Predictor, Outcome Variables, and Comparison Groups:*
(1) For the first question, the predictor variable will be `evidences` (sum of `observations`, `references`, and `vouchers`) while the outcome will be `park_name` (ordered as a factor by popularity). This will be compared across each `category_name`.

(2) For the second question the predictor variable will be `nativeness` and the outcome vairable will be `park_name` (again ordered as a factor by park popularity).

```{r}

most_visited_nps_species_data_cleaned %>% 
  filter(observations < 100) %>% # split between 2 graphs as the data was too squished
  ggplot(aes(x = observations)) +
  geom_histogram() +
  ggtitle("Distribution of Observations") + labs(subtitle = "Observations < 100")

most_visited_nps_species_data_cleaned %>% 
  filter(observations > 100) %>%
  ggplot(aes(x = observations)) +
  geom_histogram() +
  ggtitle("Distribution of Observations") + labs(subtitle = "Observations > 100")

most_visited_nps_species_data_cleaned %>%
  group_by(park_name, category_name) %>%
  summarise(sum(observations), sum(references), sum(vouchers, mean(observations), mean(references), mean(vouchers)))
```

The summary statistics help show that the amount of observations, references, and vouchers vary between parks and among the categories. The observations graphs show that there is a lot of variation in the number of observations between all the parks. It may be interesting to investigate if a particular category or park has the massive amount of observations.

# Methods
To test my questions I first would need to convert `park_name` into a factor (ordered by popularity). Then I would need to create another variable `evidences` as the sum of `observations`, `references`, and `vouchers`. It may then be helpful to create a graphic and factor by `park_name` and compare the most commonly "evidenced" categories across each park. If data follows the hyposthesized patterns, for each category `evidences` will be higher in the most popular parks. I would also want to create graphics that look at the number of different species within each park. This would be grouped by `nativeness`. If the data follows the hypothesized assotiation, parks with higher ratios of nativness will be more popular.

``` {r}
park_popularity <- c("Great Smoky Mountains", "Grand Canyon", "Zion", "Yellowstone", "Rocky Mountain", "Yosemite", "Acadia", "Grand Teton", "Joshua Tree", "Olympic", "Glacier", "Cuyahoga Valley", "Indiana Dunes", "Hot Springs", "Bryce Canyon")

#park attendance data from: 
#  Ard, A. J. (2024, June 10). These are the top 15 most visited national parks in the U.S. see which ranked no. 1. USA Today. https://www.usatoday.com/picture-gallery/travel/news/2024/06/10/what-is-the-most-visited-national-park-in-the-us-see-the-top-15/74049474007/ 

most_visited_nps_species_data_cleaned <- most_visited_nps_species_data_cleaned %>%
  mutate(
    park_name = gsub("National Park", "", park_name), # removed "National Park" in park_name since it is redundant
    park_name = factor(park_name, labels = park_popularity) # creates a factor
    ) #%>% 
  #count(park_name)       # checks order

most_visited_nps_species_data_cleaned %>%
  filter(!is.na(nativeness_tags))  # may want to combine nativeness with nativeness_tags
```
```{r}
most_visited_nps_species_data_cleaned <- most_visited_nps_species_data_cleaned%>%
  mutate(evidences = rowSums(across(c(vouchers, observations, references)), na.rm = TRUE))
```

```{r}
most_visited_nps_species_data_cleaned %>%
  count(category_name) %>%
  arrange(-n)

most_visited_nps_species_data_cleaned %>%
  group_by(category_name) %>%
  summarise(sum_evidences = sum(evidences)) %>%
  arrange(-sum_evidences)

most_visited_nps_species_data_cleaned %>%
  filter(category_name == "Vascular Plant")%>%
  group_by(park_name) %>%
  summarise(sum_evidences = sum(evidences)) %>%
  arrange(park_name)

most_visited_nps_species_data_cleaned %>%
  group_by(category_name) %>%
  summarise(sum_observations = sum(observations)) %>%
  arrange(-sum_observations)
```
```{r nativeness}


#mutate nativeness -> native/non-native
#delete nativeness_tags -> invasive, native, non-native

most_visited_nps_species_data_cleaned <- most_visited_nps_species_data_cleaned%>%
  mutate(nativeness = if_else(nativeness == "Unknown", NA, nativeness)) #nativeness = if_else(nativeness_tags == "Invasive" | nativeness_tags == "Invasive, Noxious", "Invasive", nativeness)))             #will not use invasives since not enough present with the tag


most_visited_nps_species_data_cleaned %>%
  count(nativeness, park_name) %>%
  arrange(park_name)# look at types
```
``` {r}
number_nativness_df <- 
  most_visited_nps_species_data_cleaned %>%
  group_by(park_name) %>%            # groups by park
  filter(!is.na(nativeness)) %>%     #removes NA
  select(park_name, category_name, nativeness) %>%
  count(nativeness) %>%
  mutate(native_percent = n / sum(n))

```
```{r}
number_nativness_df <- 
  number_nativness_df %>%
  mutate(group = if_else(park_name =="Great Smoky Mountains" | park_name == "Grand Canyon" | park_name == "Zion" | park_name == "Yellowstone" | park_name == "Rocky Mountain", "top5", 
                         if_else(park_name =="Glacier" | park_name == "Cuyahoga Valley" | park_name == "Indiana Dunes" | park_name == "Hot Springs" | park_name == "Bryce Canyon", "bottom5", "middle")))
```
``` {r hypothesis_nativeness}
set.seed(221)
null_dist_nativeness <- 
  number_nativness_df %>%
  filter(group != "middle", nativeness == "Non-native") %>%
 specify(response = native_percent, explanatory = group) %>%
 hypothesize(null = "independence") %>%
 generate(500, type = "bootstrap") %>%
 calculate(stat = "diff in means",
 order = c("bottom5", "top5"))
```
```{r}
ggplot(null_dist_nativeness, mapping = aes(x = stat)) +
 geom_histogram(binwidth = 0.01) +
 labs(title = "Null distribution")
```
```{r}
null_dist_nativeness %>%
 filter(stat > 0) %>%
 summarise(p_value = n()/nrow(null_dist_nativeness))
```

```{r}
most_visited_nps_species_data_cleaned %>%
  filter(!is.na(nativeness)) %>%
  ggplot(mapping = aes(y = fct_rev(park_name), #reverses order to put 1 on top
                       fill = fct_rev(nativeness))) +
  geom_bar(position = "fill") +
  labs(title = "Percent Nativness of All Species in National Parks", x = "Percent", y = "Park") +
  theme_minimal()
```

Look at top 3 categories of observations (Go back to observations not evidences)

```{r}
most_visited_nps_species_data_cleaned %>%
  filter(!park_name == "Acadia") %>%   #removes due to unusual amounts of insects
  group_by(category_name) %>%
  summarise(total_observations = sum(observations)) %>%
  arrange(-total_observations)
```
```{r}
most_visited_nps_species_data_cleaned %>%
  filter(category_name == "Bird" | category_name == "Vascular Plant" | category_name == "Mammal",
         !park_name == "Acadia") %>% #Acadia removed from graphic since the insect count was an outlier and scrunching the graph 
  group_by(park_name, category_name) %>%
  summarize(total_evidences = sum(evidences, na.rm = TRUE)) %>%
  ggplot(mapping = aes(y = fct_rev(park_name), fill = category_name, x = total_evidences)) +
  geom_col() +
  facet_wrap(~category_name, scales = "free_x") +
  theme_minimal() + 
  labs(title = "Total Evidences by Park for the Top 3 Categories", subtitle = "Excluding Acadia", x = "Count", y = "", fill = "Category") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


Hypothesis testing evidences:
```{r}
observations_df <- 
  most_visited_nps_species_data_cleaned %>%
  select(park_name, category_name, observations, evidences) %>% #selects only relevant categories
  filter(category_name == "Bird" | category_name == "Vascular Plant" | category_name == "Mammal") %>%
  group_by(category_name, park_name) %>%
  summarise(total_evidences = sum(evidences))

```
```{r}
set.seed(144)
null_dist_observations_bird <- 
  observations_df %>%
  filter(category_name == "Bird", park_name != "Acadia") %>%
   mutate(park_numeric = as.numeric(fct_inorder(park_name))) %>%
  specify(response = total_evidences, explanatory = as.numeric(park_numeric)) %>%
  hypothesize(null = "independence") %>%
  generate(reps = 500, type = "permute") %>%
  calculate(stat = "correlation")
```
```{r}
ggplot(null_dist_observations_bird, mapping = aes(x = stat)) +
 geom_histogram(binwidth = 0.1) +
 labs(title = "Null distribution")
```
```{r}
null_dist_observations_bird %>%
 filter(stat > 0) %>%
 summarise(p_value = n()/nrow(null_dist_observations_bird))
```

Mammals:
```{r}
set.seed(144)
null_dist_observations_mammals <- 
  observations_df %>%
  filter(category_name == "Mammal", park_name != "Acadia") %>%
   mutate(park_numeric = as.numeric(fct_inorder(park_name))) %>%
  specify(response = total_evidences, explanatory = as.numeric(park_numeric)) %>%
  hypothesize(null = "independence") %>%
  generate(reps = 500, type = "permute") %>%
  calculate(stat = "correlation")
```
```{r}
ggplot(null_dist_observations_mammals, mapping = aes(x = stat)) +
 geom_histogram(binwidth = 0.1) +
 labs(title = "Null distribution")
```
```{r}
null_dist_observations_mammals %>%
 filter(stat > 0) %>%
 summarise(p_value = n()/nrow(null_dist_observations_mammals))
```

Vascular Plants:
```{r}
set.seed(144)
null_dist_observations_plants <- 
  observations_df %>%
  filter(category_name == "Vascular Plant", park_name != "Acadia") %>%
   mutate(park_numeric = as.numeric(fct_inorder(park_name))) %>%
  specify(response = total_evidences, explanatory = as.numeric(park_numeric)) %>%
  hypothesize(null = "independence") %>%
  generate(reps = 500, type = "permute") %>%
  calculate(stat = "correlation")
```
```{r}
ggplot(null_dist_observations_plants, mapping = aes(x = stat)) +
 geom_histogram(binwidth = 0.1) +
 labs(title = "Null distribution")
```
```{r}
null_dist_observations_plants %>%
 filter(stat > 0) %>%
 summarise(p_value = n()/nrow(null_dist_observations_plants))
```