---
title: "Final Project Proposal"
subtitle: "National Park Species"
author: "Moriah Ruggerio"
output:
  tufte::tufte_html:
    tufte_variant: "envisioned"
    highlight: pygments
    css: ../lab.css
link-citations: true
---

```{r setup, include=FALSE}
#knitr::opts_chunk$set(eval = TRUE)

library(tidyverse)
library(skimr)
```

# Import the data:
```{r}
most_visited_nps_species_data <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2024/2024-10-08/most_visited_nps_species_data.csv')

```

``` {r tidydata}
most_visited_nps_species_data_cleaned <- most_visited_nps_species_data %>%
  janitor::clean_names()
```

Determine what variables to remove to keep it between 10-20:
```{r}

most_visited_nps_species_data_cleaned <- most_visited_nps_species_data_cleaned %>%
  filter(taxon_record_status == "Active", park_accepted == TRUE) %>%  #only includes data with active taxons, and accepted in park status
  #count(taxon_record_status)
  select(park_name:family, sci_name, common_names, occurrence:abundance, references:vouchers, te_status:s_rank) #%>% # removes empty and exluded rows to keep the variables within range
  #count(taxon_record_status, park_accepted)
  
#no data in synonyms, external_links
#all data is not sensitive (remove variable)
#combine nativeness_tabs with native & same with occurrence and occurrence_tags
#remove park_tags
#remove inactive taxons (including data rows)
#only include park accepted/remove the row and record status
```

``` {r}
write_csv(most_visited_nps_species_data_cleaned, file = "data/most_visited_nps_species_data_cleaned.csv")
```

``` {r}
most_visited_nps_species_data_cleaned %>%
  count(nativeness)
```

# Introduction

This project uses the dataset `most_visited_nps_species_data_cleaned`. The National Park Service (NPS) keeps a list of every species found within each national park. The data used in this project is originally from the National Park Species List. It can be accessed here: https://irma.nps.gov/NPSpecies/Search/SpeciesList. Due to the massive size of this dataset, it was restricted to the top 15 most popular national parks by  `f. hull`  in the `tidytuesday` git repository on September 2, 2024. This reduced dataset (`most_visited_nps_species_data`) was accessed on March 15, 2025 from https://github.com/rfordatascience/tidytuesday/blob/main/data/2024/2024-10-08/readme.md. Since this dataset originally had too many for the confines of this assignment, I restricted the scope of the data further. Only included in this projects' dataset (`most_visited_nps_species_data_cleaned`)  are species whose presence in the park has been approved by the NPS, have an active taxon (current scientific name), and a non-sensitive species status. Data on sensitive species is removed from all public access by the NPS to protect endangered or fragile species in national parks. 

The dataset has `r nrow(most_visited_nps_species_data_cleaned)` rows and `r ncol(most_visited_nps_species_data_cleaned)` variables. Each row is a specific species found in one of the parks. The variables are. (See `README.md` for variable descriptions. The intent of this research project is to examine if there is trends between biodiversity and the popularity of a park. Some specific research questions to be examined are: (1) Do more popular parks have higher observations of mammals (and other species)? (2)  Are there more native species seen in popular parks than invasives or introduced? (3) Does the amount of different species found in the park follow the popularity of the park (where more species are found in the more popular parks) the top 15 national parks.

# Data

```{r}
glimpse(most_visited_nps_species_data_cleaned)

skim(most_visited_nps_species_data_cleaned)
```