---
title: "National Park Speicess"
subtitle: "A Quick Look at Biodiversity"
author: "Moriah Ruggerio"
institute: "Houghton University"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    lib_dir: libs
    nature:
      ratio: "16:9"
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r load-packages, include = FALSE}
# Add any additional packages you need to this chunk
library(tidyverse)
library(tidymodels)
library(palmerpenguins)
library(knitr)
library(xaringanthemer)
library(skimr)
library(infer)
```

```{r setup, include=FALSE}
# For better figure resolution
knitr::opts_chunk$set(fig.retina = 3, dpi = 300, fig.width = 6, fig.asp = 0.618, out.width = "80%")
```

```{r load-data, include=FALSE}
# Load your data here
most_visited_nps_species_data_cleaned <- read_csv(file = "../data/most_visited_nps_species_data_cleaned.csv")

```

```{r}
style_xaringan(
  title_slide_background_image = "img/confetti.jpg"
)
```


class: center, middle
---

## Goal: Determine if there is any relationships between a park's popularity and biodiversity (in the top 15 most popular national parks).

--
## (1) Do more popular parks have higher evidences of birds (and other classes)?

--
## (2) Are there more native species seen in more popular parks than non-natives?

---

class: inverse, center, middle

# Setting Up the Data

---

# Evidences

- Sum of `observations`, `references`, `vouchers`

- Compilation of counts of evidences for a species

---

# Park_name

- Converted into a factor

- Ordered by park popularity

---

class: inverse, middle, center

# Do more popular parks have higher evidences of birds (and other classes)?

---

# Hypothesis

- Null Hypothesis: There is no relationship between park popularity and the number of “evidences” in each category.


- Alternative Hypothesis: There is no relationship between park popularity and the number of “evidences” in each category.

--
- Predictor = park_name (park popularity)

- Explanatory = evidences
---
``` {r park_popularity, include=FALSE}
park_popularity <- c("Great Smoky Mountains", "Grand Canyon", "Zion", "Yellowstone", "Rocky Mountain", "Yosemite", "Acadia", "Grand Teton", "Joshua Tree", "Olympic", "Glacier", "Cuyahoga Valley", "Indiana Dunes", "Hot Springs", "Bryce Canyon")

#park attendance data from: 
#  Ard, A. J. (2024, June 10). These are the top 15 most visited national parks in the U.S. see which ranked no. 1. USA Today. https://www.usatoday.com/picture-gallery/travel/news/2024/06/10/what-is-the-most-visited-national-park-in-the-us-see-the-top-15/74049474007/ 

most_visited_nps_species_data_cleaned <- most_visited_nps_species_data_cleaned %>%
  mutate(
    park_name = gsub("National Park", "", park_name), # removed "National Park" in park_name since it is redundant
    park_name = factor(park_name, levels = park_popularity) # creates a factor
    ) #%>% 
  #count(park_name)       # checks order

#most_visited_nps_species_data_cleaned %>%
  #filter(!is.na(nativeness_tags))  # may want to combine nativeness with nativeness_tags
```

```{r plot-chunk, echo=FALSE, fig.align='center', out.width='100%'}


# Code hidden with echo = FALSE
# Uses modified iris dataset from previous chunk
# Play around with height and width until you're happy with the look
most_visited_nps_species_data_cleaned %>%
  filter(category_name == "Bird" | category_name == "Vascular Plant" | category_name == "Mammal",
         !park_name == "Acadia") %>% #Acadia removed from graphic since the insect count was an outlier and scrunching the graph 
  group_by(park_name, category_name) %>%
  summarize(total_evidences = sum(evidences, na.rm = TRUE), .groups = "drop") %>%
  ggplot(mapping = aes(y = fct_rev(park_name), fill = category_name, x = total_evidences)) +
  geom_col() +
  facet_wrap(~category_name, scales = "free_x") +
  theme_minimal() + 
  labs(title = "Total Evidences by Park for the Top 3 Categories", subtitle = "Excluding Acadia", x = "Count", y = "", fill = "Category") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
 # theme options: https://ggplot2.tidyverse.org/reference/ggtheme.html
```
---
# Hypothesis Testing
```{r observations_df, include=FALSE}
observations_df <- 
  most_visited_nps_species_data_cleaned %>%
  select(park_name, category_name, observations, evidences) %>% #selects only relevant categories
  filter(category_name == "Bird" | category_name == "Vascular Plant" | category_name == "Mammal") %>%
  group_by(category_name, park_name) %>%
  summarise(total_evidences = sum(evidences))

```
.pull-left[
**Birds:**
```{r bird-p, echo=FALSE}
set.seed(144)
null_dist_observations_bird <- 
  observations_df %>%
  filter(category_name == "Bird", park_name != "Acadia") %>%
   mutate(park_numeric = as.numeric(fct_inorder(park_name))) %>%
  specify(response = total_evidences, explanatory = park_numeric) %>%
  hypothesize(null = "independence") %>%
  generate(reps = 500, type = "permute") %>%
  calculate(stat = "correlation")

null_dist_observations_bird %>%
 filter(stat > 0) %>%
 summarise(p_value = n()/nrow(null_dist_observations_bird))
```

**Mammals:**
``` {r mammal-p, echo=FALSE}
set.seed(144)
null_dist_observations_mammals <- 
  observations_df %>%
  filter(category_name == "Mammal", park_name != "Acadia") %>%
   mutate(park_numeric = as.numeric(fct_inorder(park_name))) %>%
  specify(response = total_evidences, explanatory = park_numeric) %>%
  hypothesize(null = "independence") %>%
  generate(reps = 500, type = "permute") %>%
  calculate(stat = "correlation")

total_n <- nrow(null_dist_observations_mammals)
null_dist_observations_mammals %>%
 filter(stat > 0) %>%
 summarise(p_value = n()/total_n)
```
]

.pull-right[
**Plants:**
``` {r plants-p, echo=FALSE}
set.seed(144)
null_dist_observations_plants <- 
  observations_df %>%
  filter(category_name == "Vascular Plant", park_name != "Acadia") %>%
   mutate(park_numeric = as.numeric(fct_inorder(park_name))) %>%
  specify(response = total_evidences, explanatory = park_numeric) %>%
  hypothesize(null = "independence") %>%
  generate(reps = 500, type = "permute") %>%
  calculate(stat = "correlation")

total_n <- nrow(null_dist_observations_plants)
null_dist_observations_plants %>%
 filter(stat > 0) %>%
 summarise(p_value = n()/total_n)
```
]

---

class: inverse, middle, center

# Are there more native species (as compared to non-natives) seen in more popular parks than less popular? 

---



---
# Layouts

You can use plain text

- or bullet points

.pull-left[
or text in two columns $^*$
]
.pull-right[
- like
- this
]

.footnote[
[*] And add footnotes
]

---

# Code



---

# Plots

```{r recode-species, echo = FALSE}
# In this chunk I'm doing a bunch of analysis that I don't want to present 
# in my slides. But I need the resulting data frame for a plot I want to present.
iris_modified <- iris %>%
  mutate(Species = fct_other(Species, keep = "setosa"))
```

```{r plot-iris, echo = FALSE}
# Code hidden with echo = FALSE
# Uses modified iris dataset from previous chunk
# Play around with height and width until you're happy with the look
ggplot(data = iris_modified, mapping = aes(x = Sepal.Width, y = Sepal.Length, color = Species)) +
  geom_point() + 
  theme_minimal() # theme options: https://ggplot2.tidyverse.org/reference/ggtheme.html
```

---

## Plot and text

.pull-left[
- Some text
- goes here
]
.pull-right[
```{r warning=FALSE, out.width="100%", fig.width=4, echo=FALSE}
# see how I changed out.width and fig.width from defaults
# to make the figure bigger
ggplot(penguins, aes(x = bill_length_mm, y = species, color = species)) +
  geom_boxplot() +
  theme_minimal()
```
]

---

# Tables

If you want to generate a table, make sure it is in the HTML format (instead of Markdown or other formats), e.g.,

```{r iris-table, echo = FALSE}
kable(head(iris), format = "html")
```

---

# Images

```{r castle, echo = FALSE, out.width = "60%", fig.align = "center", fig.cap = "Image credit: Photo by Jörg Angeli on Unsplash."}
include_graphics("https://images.unsplash.com/photo-1535448033526-c0e85c9e6968?ixlib=rb-1.2.1&ixid=eyJhcHBfaWQiOjEyMDd9&auto=format&fit=crop&w=1650&q=80")
```

Or you can also include a full page image. See next slide.

---

background-image: url(https://images.unsplash.com/photo-1535448033526-c0e85c9e6968?ixlib=rb-1.2.1&ixid=eyJhcHBfaWQiOjEyMDd9&auto=format&fit=crop&w=1650&q=80)

---

# Math Expressions

You can write LaTeX math expressions inside a pair of dollar signs, e.g. &#36;\alpha+\beta$ renders $\alpha+\beta$. You can use the display style with double dollar signs:

```
$$\bar{X}=\frac{1}{n}\sum_{i=1}^nX_i$$
```

$$\bar{X}=\frac{1}{n}\sum_{i=1}^nX_i$$

Limitations:

1. The source code of a LaTeX math expression must be in one line, unless it is inside a pair of double dollar signs, in which case the starting `$$` must appear in the very beginning of a line, followed immediately by a non-space character, and the ending `$$` must be at the end of a line, led by a non-space character;

1. There should not be spaces after the opening `$` or before the closing `$`.

1. Math does not work on the title slide (see [#61](https://github.com/yihui/xaringan/issues/61) for a workaround).

---

# Feeling adventurous?

- Want to find out more about `xaringan`? See https://slides.yihui.name/xaringan/#1.

- You are welcomed to use the default styling of the slides. In fact, that's what I expect majority of you will do. You will differentiate yourself with the content of your presentation.

- But some of you might want to play around with slide styling. The 
`xaringanthemer` provides some solutions for this that: https://pkg.garrickadenbuie.com/xaringanthemer.

- And if you want more bells and whistles, there is also `xaringanExtra`: https://pkg.garrickadenbuie.com/xaringanExtra.
